---
title: "Permutation tests for statistical significance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Permutation tests for statistical significance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: '`r system.file("REFERENCES.bib", package="transclust")`'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 320
)
```

```{r setup}
library(transclust)
library(ape)
library(ggplot2)
```

This vignette provides an overview of the permutation tests for statistical significance in the `transclust` package. The package includes functions for performing permutation tests to assess the statistical significance of clustering results. These tests can be used to determine whether the observed clustering is significantly different from what would be expected by chance.

## Loading and preparing data

These steps are the same as those in the "Introduction to transclust" vignette. The `transclust` package comes with a sample dataset that can be used for demonstration purposes. The dataset contains epidemiological information for a set of isolates. To load the dataset, we can use the following command:

```{r load_data}
load(system.file("extdata", "example.Rdata", package = "transclust"))
```

We first read in the sequence file which has been prepared using the procedure described in @Hawken2022. We can read in the sequence file using `ape`'s `read.dna` function:

```{r read_seq}
dna_aln <- read.dna(system.file("extdata", "example.fasta", package = "transclust"), format = "fasta")
```

Now that the sequence file is loaded, we can extract the variable positions in the alignment. The `dna_var_pos` variable contains a logical vector indicating which columns in the alignment are variable. The `dna_pt_labels` variable contains the labels for the sequences in the alignment, and `trace_mat` is a matrix containing the trace data for the isolates. The valid set of labels is determined by checking which labels in the `dna_pt_labels` variable are present in the `trace_mat` matrix.

```{r var_pos}
# Get all variable positions in the alignment
var_pos <- apply(dna_aln, 2, function(x) sum(x == x[1]) < nrow(dna_aln))
# Only keep those labels that are in the trace matrix
valid_labels <- dna_pt_labels[labels(dna_aln)] %in% colnames(trace_mat)
```

We can then subset the alignment to include only the variable positions from the isolates with valid labels:

```{r subset_var}
dna_var <- dna_aln[valid_labels, var_pos]
```

We then use the helper function `get_snp_dist_matrix()` to calculate the SNP distance matrix for these isolates:

```{r snp_dist}
snp_dist <- get_snp_dist_matrix(dna_var)
```

The `snp_dist` variable contains the SNP distance matrix, which is a square matrix where each entry represents the number of SNP differences between two isolates. The diagonal entries are all zero, as they represent the distance between an isolate and itself.

## Clustering

`transclust` offers two methods to determine transmission clusters: these include a hard SNP-threshold based approach and a threshold-free shared-variant based algorithm as described in @Hawken2022. In this vignette, we will focus on the latter method for statistical analysis. We first create a phylogenetic tree using the SNP distance matrix and use that along with the epidemiological data to get the transmission clusters:

```{r phylo_tree, message = FALSE}
phylo_tree_pars <- get_phylo_tree(dna_var, snp_dist, "pars")
clusters <- get_tn_clusters_sv_index(
    dna_var, snp_dist, ip_seqs_3days, ip_seqs,
    dna_pt_labels, dates, phylo_tree_pars
)
```

This gives us the clusters that we can now further use for statistical analysis.


## Permutation tests

The `transclust` package includes a function that performs permutation tests to assess the statistical significance of clustering results. A list of several important prooperties of the clusters are computed over the observed data and the permuted data. The permutation test is performed by randomly permuting the labels of the isolates and recalculating the clustering results. This process is repeated a specified number of times to generate a distribution of clustering results under the null hypothesis. The observed clustering results are then compared to this distribution to determine whether they are significantly different from what would be expected by chance.

For the tests to be informative, the clusters have to be permuted with certain constraints. We can divide our patients into two "classes": index and convert. Index patients are those who are the first in a cluster, and convert patients are those who are infected by an index patient. The permutation tests are performed by randomly permuting the labels of the isolates while not changing the proportions of index and convert patients in the clusters. This is done to ensure that the permutation tests are meaningful and take into account the structure of the data.

Because the permutation tests are computationally intensive and can take time, for better performance they can be run in parallel on multiple cores on *nix systems (Windows does not support parallel execution in this manner). However, since this is a vignette that has to on all platforms, we will only run the test on a single core. The `cluster_property_perm_test()` function takes the following arguments:

```{r perm_test, warning = FALSE}
nperm <- 500
perm_props <- cluster_property_perm_test(
    clusters, trace_mat, dna_pt_labels, ip_seqs, ip_seqs_3days, dates, snp_dist,
    nperm = nperm, floor_trace = floor_mat, room_trace = room_mat, num_cores = 1
)
```

The `perm_props` object contains the results of the permutation tests, including the observed and permuted values for each property. The `cluster_property_perm_test()` relies on the `cluster_properties()` function to calculate the properties of the clusters along with some additional properties that are calculated after the permutation tests. How do we know which properties? We can see what the properties are by looking at the `perm_props` object:

```{r perm_props}
colnames(perm_props)
```

Lets choose to look at one of these, the `"Median Max_genetic_distance"`. This is just the median of the maximum genetic distance between isolates in a cluster. We can plot the observed value against the permuted values to see how it compares:

```{r plot_perm, warning = FALSE}
prop_name <- colnames(perm_props)[11]

perm_vals <- perm_props[1:nperm, prop_name]
observed_val <- perm_props[nperm + 1, prop_name]

upper_p <- (sum(perm_vals <= observed_val) + 1) / (nperm + 1)
lower_p <- (sum(perm_vals >= observed_val) + 1) / (nperm + 1)
prop_pval <- min(upper_p, lower_p)

ggplot(data = data.frame(x = perm_vals), aes(x = x)) +
    geom_histogram(aes(y = after_stat(count)), bins = 10, fill = "lightblue", color = "black") +
    geom_point(aes(x = observed_val, y = 0), shape = 23, size = 5, fill = "red") +
    annotate("text", x = observed_val, y = 0.5, label = "Observed", vjust = -1) +
    labs(
        title = paste0("Permutation tests for ", prop_name, "\n p-value: ", round(prop_pval, 3)),
        x = prop_name,
        y = "Count"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
```

We can see that the observed value is _significantly_ lower than the permuted values, indicating that our observed clusters have mostly closely related isolates. The p-value is calculated as the proportion of permuted values that are less than or equal to the observed value. A low p-value indicates that the observed value is significantly different from what would be expected by chance. The p-value we see here is 0.01, which is lower than the commonly used significance level of 0.05.

What does this mean in this case? Well, the maximum genetic distance between isolates in a cluster is significantly lower than what would be expected by chance. This suggests that the clusters are made up of closely related isolates, which is consistent with the idea that they are part of the same transmission chain. This is an important finding, as it suggests that the clusters are not just random groups of isolates, but rather represent real transmission events.

Let's look at another example, this time that is not statistically significant. We can look at the `"Median Number_of_converts_after_index"` property, which is the median number of converts after the index case in a cluster. This property is not expected to be significantly different from what would  be expected by chance, as it is not directly related to the clustering results and is in fact a property of the underlying data. We can plot the observed value against the permuted values to see how it compares:

```{r plot_perm2, warning = FALSE}
prop_name <- colnames(perm_props)[14]

perm_vals <- perm_props[1:nperm, prop_name]
observed_val <- perm_props[nperm + 1, prop_name]

upper_p <- (sum(perm_vals <= observed_val) + 1) / (nperm + 1)
lower_p <- (sum(perm_vals >= observed_val) + 1) / (nperm + 1)
prop_pval <- min(upper_p, lower_p)

ggplot(data = data.frame(x = perm_vals), aes(x = x)) +
    geom_histogram(aes(y = after_stat(count)), bins = 10, fill = "lightblue", color = "black") +
    geom_point(aes(x = observed_val, y = 0), shape = 23, size = 5, fill = "red") +
    annotate("text", x = observed_val, y = 0.5, label = "Observed", vjust = -1) +
    labs(
        title = paste0("Permutation tests for ", prop_name, "\n p-value: ", round(prop_pval, 3)),
        x = prop_name,
        y = "Count"
    ) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5))
```

Here we can see the p-value is very close to 1, indicating that the observed value is not significantly different from what would be expected by chance. This is because our permutations are designed to keep similar proportions of index and convert patients in the clusters, and so we would not expect this property to be significantly different from what would be expected by chance.

Thus the permutation tests provide a useful way to assess the statistical significance of clustering results, and can help to identify properties that are significantly different from what would be expected by chance, helping us to understand signatures of transmission in the data.

## References

<div id="refs"></div>
