---
title: "Introduction to transclust"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to transclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: '`r system.file("REFERENCES.bib", package="transclust")`'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    dpi = 320
)
```

This vignette provides a brief introduction to the `transclust` package, which is designed for analysis of transmission clusters using a combination of whole genome sequencing (WGS) data and epidemiological information. The package includes functions for clustering, visualization, and statistical analysis of transmission clusters. To get started, we will first load the package. We will also load the `ape` package, which is used for phylogenetic analysis with the package and in the vignette, and `tidyr`, `ggplot2` and `paletteer` for data manipulation and visualization.

```{r setup}
library(transclust)
library(ape)
library(tidyr)
library(ggplot2)
library(ggalign)
library(paletteer)
```

## Loading and preparing data

The `transclust` package comes with a sample dataset that can be used for demonstration purposes. The dataset contains epidemiological information for a set of isolates. To load the dataset, we can use the following command:

```{r load_data}
load(system.file("extdata", "example.Rdata", package = "transclust"))
```

We first read in the sequence file which has been prepared using the procedure described in @Hawken2022. We can read in the sequence file using `ape`'s `read.dna` function:

```{r read_seq}
dna_aln <- read.dna(system.file("extdata", "example.fasta", package = "transclust"), format = "fasta")
```

Now that the sequence file is loaded, we can extract the variable positions in the alignment. The `dna_var_pos` variable contains a logical vector indicating which columns in the alignment are variable. The `dna_pt_labels` variable contains the labels for the sequences in the alignment, and `trace_mat` is a matrix containing the trace data for the isolates. The valid set of labels is determined by checking which labels in the `dna_pt_labels` variable are present in the `trace_mat` matrix.

```{r var_pos}
# Get all variable positions in the alignment
var_pos <- apply(dna_aln, 2, function(x) sum(x == x[1]) < nrow(dna_aln))
# Only keep those labels that are in the trace matrix
valid_labels <- dna_pt_labels[labels(dna_aln)] %in% colnames(trace_mat)
```

We can then subset the alignment to include only the variable positions from the isolates with valid labels:

```{r subset_var}
dna_var <- dna_aln[valid_labels, var_pos]
```

We then use the helper function `get_snp_dist_matrix()` to calculate the SNP distance matrix for these isolates:

```{r snp_dist}
snp_dist <- get_snp_dist_matrix(dna_var)
```

The `snp_dist` variable contains the SNP distance matrix, which is a square matrix where each entry represents the number of SNP differences between two isolates. The diagonal entries are all zero, as they represent the distance between an isolate and itself.

## Clustering isolates using a hard SNP threshold

Now that we've loaded our data, we can start clustering the isolates. First, we want to use a hard SNP threshold to cluster the isolates.

The `get_tn_clusters_snp_thresh()` function takes the SNP distance matrix and a threshold value as input and returns a list of clusters. We use a threshold of 10 SNPs for this example:

```{r snp_clusters}
clusters_snp <- get_tn_clusters_snp_thresh(snp_dist, 10)
```

We can then visualize the clusters on a phylogenetic tree using the `plot_clusters_phylo()` function. But for this, we first need to generate a phylogenetic tree. We can do this using the `get_phylo_tree()` function. We use the maximum parsimony method for this example:

```{r phylo_tree_pars, results = "hide"}
# Generate a parsimony tree
phylo_tree <- get_phylo_tree(dna_var, snp_dist, "pars")
```

```{r plot_snp_clusters, fig.width=6, fig.height=10}
plot_clusters_phylo(phylo_tree, clusters_snp)
```

But this isn't the only SNP threshold approach. This method tries to preserve clusters to be within the phylogenetic tree structure, but we can also cluster the isolates just based on the SNP distance matrix with no regard for the 

## Clustering isolates using a threshold-free approach

In addition to the hard SNP threshold approach, we can also use a threshold-free approach to cluster the isolates. The package implements the method described in @Hawken2022 in the `get_tn_clusters_sv_index()` function. The method uses a threshold-free approach to identify clusters based on the structure of the phylogenetic tree. We use the same parsimony tree as before for this example. This function takes in a few more inputs than the SNP threshold method:

```{r sv_clusters, message = FALSE}
clusters_sv <- get_tn_clusters_sv_index(
    dna_var, snp_dist, ip_seqs_3days, ip_seqs,
    dna_pt_labels, dates, phylo_tree
)
```

The `clusters_sv` object contains the clusters identified using the threshold-free approach. The plot shows the phylogenetic tree with the clusters highlighted.

```{r plot_sv_clusters, fig.width=6, fig.height=10}
plot_clusters_phylo(phylo_tree, clusters_sv)
```

## Comparing clusters

These two trees look quite different. We can compare the clusters generated using the two methods using the `compare_clusters()` function. This function generates a heatmap showing the overlap between the clusters by indicating the proportion of isolates in each cluster that are also present in the clusters generated by the other method. Before we can use this function, we need to remove any singleton clusters, which are clusters that contain only one isolate – these are not very useful for comparison:

```{r compare_clusters, fig.width=10, fig.height=10}
# remove singleton clusters from both clustering methods
clusters_snp <- remove_singleton_clusters(clusters_snp)
clusters_sv <- remove_singleton_clusters(clusters_sv)
# compare the clusters
compare_clusters(clusters_snp, clusters_sv) +
    # change the color palette
    scale_fill_paletteer_c("grDevices::Mint", direction = -1) +
    # add a title
    ggtitle("Comparison of clusters using SNP threshold and threshold-free methods") +
    # add x and y axis labels
    xlab("Threshold-free clusters") +
    ylab("SNP threshold clusters") +
    # center the title and rotate x axis labels
    theme(
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1)
    )
```

## Some statistical analysis of the clustering results

The `transclust` package also provides functions for statistical analysis of the clustering results. For example, we can use the `intra_cluster_genetic_var_analysis()` function to calculate the genetic variation within each cluster:

```{r cluster_stats}
mut_var_df <- intra_cluster_genetic_var_analysis(clusters_sv, dna_aln, var_pos)
```

This dataframe has columns for the total number of variable sites, and also the proportions of each of the six types of base substitutions present in each cluster. But this isn't very intuitive to just look at, so we can plot this. First, we drop any possible NA values. For this plot, we only want to compare relative frequencies of the different types of base substitutions, so we also exclude the first column - the total number of variable sites. We further subset the data frame to only include the first few rows for clarity in the plot:

```{r drop_na_subset}
mut_var_df <- head(na.omit(mut_var_df)[, -1])
```

Then we can plot the proportions of each type of base substitution in each cluster. For this, the `ggplot2` package is used to create a stacked bar plot:

```{r plot_cluster_stats, fig.width=10, fig.height=6}
# Add column for cluster using rownames before pivoting
mut_var_df$cluster <- row.names(mut_var_df)

# Convert the data frame to long format for ggplot
mut_var_df_long <- pivot_longer(
    mut_var_df,
    cols = -cluster,
    names_to = "mutation_type",
    values_to = "proportion"
)

# Create the stacked bar plot
ggplot(mut_var_df_long, aes(x = cluster, y = proportion, fill = mutation_type)) +
    geom_bar(stat = "identity") +
    labs(x = "Cluster", y = "Proportion", fill = "Mutation Type") +
    scale_fill_paletteer_d("ggsci::lanonc_lancet")
```

We can clearly see very specific mutational signatures emerge in the clusters when compared to the population as a whole.

## References

<div id="refs"></div>
