---
title: "Input data format"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

`hospitraceR` works by fusing two kinds of evidence: the genomes of the isolates and the
epidemiological record of the patients they came from. Rather than one large object, the package
takes each piece as a separate, lightweight R object, and different functions ask for the pieces
they need. This article catalogs every input, the exact form it should take, and which functions
consume it. An example dataset is used throughout as a concrete reference.

```{r load}
library(hospitraceR)
library(ape)

# An example dataset of isolates and their patient epidemiological metadata.
extdata_dir <- file.path("..", "..", "inst", "extdata")
load(file.path(extdata_dir, "example.RData"))
dna_aln <- read.dna(file.path(extdata_dir, "example.fasta"), format = "fasta")
```

Two identifiers run through everything and are worth fixing up front:

- a **sequence ID** (also called an isolate ID) names a single sequenced isolate — one row of the
  alignment;
- a **patient ID** names a patient, who may have contributed several isolates.

Most per-sequence inputs are **named vectors** whose *names* are sequence IDs. The mapping from
sequence to patient is what ties the genomic and epidemiological sides together.

# The sequence alignment

The genomic input is a multiple sequence alignment of class `DNAbin` (as produced by
`ape::read.dna()`), with one row per isolate and the sequence IDs as row labels.

```{r aln}
dna_aln
head(labels(dna_aln))
```

By convention the **first sequence is an outgroup** — in the example, the `KPNIH1` reference genome.
`get_tn_clusters_sv_index()` and the tree it consumes both treat the first alignment sequence (and
the first tip of the tree) as that outgroup, so it should stay in position one. From the alignment,
you can derive the SNP distance matrix with `get_snp_dist_matrix()` and a phylogenetic tree with
`get_phylo_tree()`. Any other sources for SNP distances or trees should also be fine, as long as 
the sequence IDs match up.

Both clustering methods then work from variable sites only (see the
[introductory article](hospitraceR.html)).

# Per-sequence epidemiological vectors

## Sequence-to-patient map

A named integer (or character) vector mapping each sequence ID (the names) to its patient ID (the
values). Functions take it as the `seq2pt` argument; in the example dataset it is `dna_pt_labels`.

```{r seq2pt}
head(dna_pt_labels)
```

Here sequence IDs `1`, `2` and `3` all belong to patient `6` — three isolates from one patient.

## Collection dates

A named numeric vector of collection dates, one per sequence, named by sequence ID. Dates are
expressed as integer day numbers on a common timeline (day `0` is the start of the study window),
*not* as `Date` objects, because the location traces below are indexed by the same day numbers.

```{r dates}
head(dates)
```

## Admission status

Two vectors of sequence IDs encode who was already carrying the organism at admission:

- `adm_seqs` — sequences from patients who were **positive at intake** (admission-positive at the
  time of their first screen);
- `adm_pos_pt_seqs` — **all** sequences from admission-positive patients, whether collected at
  intake or later. This is a superset of `adm_seqs`: it adds the later isolates of those same
  patients.

```{r adm}
length(adm_seqs)
length(adm_pos_pt_seqs)
all(adm_seqs %in% adm_pos_pt_seqs)
```

The distinction matters for clustering: a patient positive at intake cannot have acquired the
organism in the facility, so admission-positive sequences anchor clusters as potential sources
rather than converts.

# Surveillance cultures

`surv_df` is a data frame of surveillance results, one row per culture, with four columns:

| Column | Meaning |
|--------|---------|
| `patient_id` | the patient screened |
| `genome_id` | the sequence ID, if the culture was sequenced |
| `surv_date` | the screen date, on the same day-number timeline as `dates` |
| `result` | `1` for a positive culture, `0` for a negative one |

```{r surv}
head(surv_df)
```

This is what lets the package tell a genuine conversion (a previous-negative patient who later turns
positive) from a patient who was already colonized: `get_isolate_lookup()` reads `surv_df` to derive
each isolate's `prev_surv` and `prev_surv_neg` flags (see the
[overlap analysis article](overlap_analysis.html)).

# Location traces

The patients' movements are given as **trace matrices**: patients in rows, days in columns. A cell
records where a patient was on a given day, as an integer location code, with `0` meaning the
patient was not present (e.g. discharged). The column names are the day numbers, on the same
timeline as `dates` and `surv_date`.

The package uses three traces at increasing spatial resolution, all the same shape:

- `facility_trace` — A `0`/`1` matrix, present or not;
- `floor_trace` — which floor;
- `room_trace` — which room.

```{r trace}
dim(floor_trace)
floor_trace[1:5, 1:8]
sort(unique(as.vector(floor_trace))) # 0 = absent; 1-6 = floors
```

Two patients "overlap" on a day when their trace cells hold the same positive code (the same floor,
say). The overlap and permutation-test functions
(`isolate_isolate_overlap()`, `cluster_overlap_perm_test()`, ...) take whichever trace matrix
matches the resolution you want to test at.

# Which function needs what

Putting it together, the inputs flow into the package as follows:

| Input (example object) | Form | Used by |
|------------------------|------|---------|
| Alignment (`dna_aln`) | `DNAbin`, first row = outgroup | `get_snp_dist_matrix()`, `get_phylo_tree()`, `get_tn_clusters_sv_index()` |
| SNP distances | numeric matrix | `get_tn_clusters_snp_thresh()`, `get_tn_clusters_sv_index()`, distance metrics |
| Tree | `phylo`, first tip = outgroup | `get_tn_clusters_sv_index()`, `plot_trace_phylo_tree()` |
| Sequence-to-patient (`dna_pt_labels`) | named vector, seq ID -> patient ID | clustering, `get_isolate_lookup()`, permutation test |
| Dates (`dates`) | named numeric, seq ID -> day | clustering, `get_isolate_lookup()` |
| Admission status (`adm_seqs`, `adm_pos_pt_seqs`) | sequence-ID vectors | `get_tn_clusters_sv_index()`, `get_isolate_lookup()`, permutation test |
| Surveillance (`surv_df`) | data frame (4 columns above) | `get_isolate_lookup()`, categorization, permutation test |
| Traces (`facility_trace`, ...) | patient-by-day matrices | overlap functions, permutation test |

Most analyses begin by passing the clustering inputs to `get_tn_clusters_sv_index()`, then feeding
the resulting clusters and the remaining inputs to `get_isolate_lookup()`, whose lookup table the
downstream overlap, categorization, and statistics functions all consume.
